An integrated open population distance sampling approach for modelling age-structured populations

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Added discussion section

Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo. Overall, the IDSM was able to corretly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. Estimates of population sizes / densities were also adequate, and are presented in the Supplementary Materials (see folder SimCheck_byDataSet).

@@

-   In general, the model does a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known. For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.

@@

-   The fact that this formulation allow us to separately estimate the underlying demographic rates (survival and recruitment) is a clear improvement compared to previous open population DS models, as this allow us both to obtain estimates of the rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$) .
-   Nevertheless, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate. This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). However, the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small. \[A few sentences about fading rodent cycles, and perhaps a weakened link to ptarmigan dynamics\]. \[Also: a time-series of 15 years is still relatively short, so it is to be expected that we have low statistical power to detect temporal covariate effects --\> more years of data, or space-for-time substitution, will help)
-   In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the version presented in the results section, we made the assumption that these birds were in fact juveniles. In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix). As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected. Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density. Compared to previous studies on ptarmigan, we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
-   The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates. Another one might be data that allows to distinguish between true and apparent survival<!-- or is that accounted for in the telemetry data, i.e. are the individuals that left hte study area censored? -->. It is also possible to extent the model with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates. \[See my point above on limited statistical power with 15 years. Maybe could be good to link to the space-for-time substitution idea here and "tease" the extension to multi-area since the monitoring programme, at least for line transects, covers many more areas).

>>>>

Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo. Overall, the IDSM was able to correctly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. Estimates of population sizes / densities were also adequate, and are presented in the Supplementary Materials (see folder SimCheck_byDataSet).

@@

We developed a model that use line transect distance sampling survey data combined with data from marked individuals to estimate population abundance, survival probabilities and recruitment rates in an open population setting. We first used simulated data to examine models ability to recover the underlying parameters when they were known. Then we fitted the model to data from an ongoing field study on willow ptarmigan in Norway. In general, the model did a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known. For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters. When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from distance sampling data.

Open population formulations of the distance sampling model has previously been presented (refs), and applied to various ecological systems (refs). Our model extends those previous applications by formulating the underlying process model as a stage-structured matrix model where the matrix elements are represented by annual survival probabilities and recruitment rates. While such models have been available for a range of other modelling frameworks previously (in particular), this is to our knowledge the first explicit use of such a state model in a distance sampling framework. Such a formulation allow us to both to obtain estimates of the underlying demographic rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$). As an example of such environmental variation, we investigated the effects of small rodent abundance on the recruitment rates. The fact that we were able to successfully recover the covariate effect when the model was applied to simulated data, suggest that in fact the model is able to estimate the effect when there is a clear signal in the data. When applied to real data, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate. This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). We speculate that there are (at least) three potential explenations to this result. First, for our specific case study the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small. It is well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception. Lack of peak rodent years in the time series to which we fitted the model. In addition, previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); we did not fit any weather covariates to the model so there is a possibility that strong effects of spring conditions in certain years have masked any remaining effects of small rodent abundance. Finally, the data set we used is relative short (15 years), leaving us with somewhat limited statistical power. We therefore do not consider our a particularily strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. In future applications, a stronger test based on more years of data and data from additional areas is encouraged. In e.g. [@Bowler2020] we used data from the same sampling program but from more areas, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. Our new model allow us to investigate to which extent similar results emerge when we can link the covariate effect to the actual demographic rate and not only to the population growth rate.

One potential source of bias in our model framework is related to the way that we use the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and thsi create a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the version presented in the results section, we made the assumption that these birds were in fact juveniles. In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix). As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected. Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density. Compared to previous studies on ptarmigan (see e.g. [@Israelsen2020][@Sandercock2011]), we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification. In future developments of the model, explicit accounting for such misclassification will be considered.

@@

In summary, we conclude that the model presented here is useful in many ecological systems where the sampling survey data include information about the age-ratio of the observed individuals. Following the integrated modelling philosophy, it also allow for the integration of auxiliary data. In our application here, we integrated data from marked birds by formulating an explicit (known-fate) mark-recapture model that explicitly supported the estimation of survival probabilities in the state process. The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates. Moreover, the model can be extended with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates. Finally, as the underlying model structure is based on a hierarchical model the framework readily support multi-area extensions, where the covariate-effects are efficiently estimated by time-for-space substitution.
